Inference using context-dependent models of mutation via bounds on information propagation

### 2021-12-04 writeups/context-dependence.tex
.

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\calP}{\mathcal{P}}  % set of selection pairs
\newcommand{\calM}{\mathcal{M}}  % set of potential mutations
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\nS}{\mbox{S}} % C or G
\newcommand{\nW}{\mbox{W}} % A or T
\newcommand{\nR}{\mbox{R}} % purine, A or G
\newcommand{\nY}{\mbox{Y}} % pyramidine, C or T
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}
\newcommand{\like}{\mathcal L}
\newcommand{\loglike}{\ell}
\newcommand{\alike}{\widetilde{\like}}
\newcommand{\aloglike}{\widetilde{\loglike}}

@@

Early models of DNA sequence mutation
assumed that each nucleotide was equally likely to mutate
into any other nucleotide \citep{jukes1969evolution}.
These were followed by others that incorporated differences 
in the mutation process, such as higher transition to transversion ratio
\citep{hasegawa1985dating,felsenstein1996hidden}.
Although the rates of mutation of a nucleotide
are known to depend on not only the nucleotide itself
but also nearby sequence context \citep{seplyarskiy2021population},
it is difficult to take this context into account.

This situation from biological sequence analysis
is an example of a more commonly encountered class of models:
a lattice of sites,
with each site taking one of a finite collection of possible states,
and whose stochastic, temporal evolution is Markov and governed by a set of local rules.
For instance, at each site may sit
a nucelotide whose mutation rate depends on physical properties determined by the nearby DNA sequence;
a cell whose infection status depends on the state of its neighbors;
or a particle whose spin is perturbed by external noise to a state whose distribution depends on the local energy configuration.
Practical use of such models often requires
inferring transition rules
based on observations of the system at several time points,
or of several states evolved along a tree from a single starting point.
This paper concerns scalable methods for doing inference under such models using observations.

@@

In the context of genomics,
it is well-known that certain short nucleotide sequences
are in many organisms much more, or less, abundant than expected by chance \citep{burge1992underrepresentation},
due to the combined effects of context dependence of the nucleotide mutation process,
selective constraints on the function of the sequence,
and other processes such as biased gene conversion \citep{duret2009biased,arbeithuber2015crossovers}.
Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
is highly context dependent \citep{schaibley2013influence,gao2019overlooked,rodriguezgalindo2020germline}
and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution,herediagenestar2020extreme},
which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
In general, substitution rates depend on sequence context due to the interaction of a great many factors including:
(a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{goodman2002errorprone} or crossing over in meiosis \citep{arbeithuber2015crossovers});
(b) activity levels and structural properties of the enzymes responsible for repair
(e.g., Y-family polymerases \citep{goodman2013translesion,sale2012yfamily});
and (c) any functional effects of an error (e.g., lethal mutations will not be seen in a population census).
We are concerned with effects that are \emph{homogeneous} across the sequence,
so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.
Some of these effects -- particularly, activity levels of different enzymes in the germline --
will likely vary over evolutionary time,
and so comparisons between widely separated species will identify time-averaged mutation rates.
Studies have used a variety of exploratory techniques to identify sets of mutational patterns
co-occurring in comparisons (a) between different cancer types \citep{alexandrov2013signatures,shiraishi2015simple} and
(b) between different human populations \citep{harris2015evidence,harris2016rapid,mathieson2017differences},
to disentangle distinct signals presumably coming from the action of distinct sources of mutation.

@@

To formalize this notion of local dependence,
first define a \textit{pattern} to mean a contiguous sequence of states;
let $|u|$ denote the length of the pattern $u$.
Given $x \in \calS$,
let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
% Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
% This is natural in many situations, but in others we might want to other boundary conditions,
% modeling extant, but unobserved, neighbors.
% We will mention this when necessary.
The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$,
written as ``$u \to v$ at rate $\mu$''.
The set of all transition triples is denoted $\calT$.

\begin{example}[TASEP]
  The \emph{Totally Asymmetric Simple Exclusion Process} labels each site as ``empty'' or ``occupied'' (0 or 1, respectively),
  and says that each occupied site, independently at rate $\lambda$, checks the site to the right to see if it is empty,
  and if it is, moves there.
  This process therefore has only one transition triple:

@@

  \noindent
  This only has one parameter.
  If we allow particles to exit from the right, and new particles to enter from the left
  then from observing only starting and ending configurations
  we cannot tell with certainty which particles have moved where,
  so it is not obvious how to estimate the speed.
  However, summing over possible particle movements would allow us to
  compute the likelihood of a given configuration change,
  and so estimate the speed by maximum likelihood
  given an initial sequence, a final sequence and an elapsed time.

@@

To allow more compact specification of models,
we also introduce a ``potential'':
suppose that there is a collection of patterns $\{p_i\}$
such that each occurrence of $p_i$ adds a quantity $e_i$ to the total ``energy'' of a sequence,
and that the rate at which each possible change occurs is modulated by a function of the energy difference the change would produce.
This is natural if that transition triples describe \emph{proposed} changes,
and that the probability a proposed change actually occurs depends on how much it affects the energy.

@@

\begin{example}[Genomic GC content]
    A sequence of genome can be written using A, C, G, and T;
    in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
    Furthermore, it is a well-known observation that in many species,
    adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
    than either single nucleotide change under the independent model.
    Embellishing the single-nucleotide model with this additional rate results in the model defined by
    \begin{center}
      \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
        $x$  &  $y$  &  $m_{xy}$ & for $x, y \in \{\nA,\nC,\nG,\nT\}$ and $x \neq y$ \\
        \nC\nG   &  \nT\nG   &  $\gamma$ & \\
        \nC\nG   &  \nC\nA   &  $\gamma$ &
      \end{tabular},
    \end{center}
    where $\gamma$ is the additional CpG rate above the base mutation rate.
    Note that for a given pair of sequences,
    a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a $\nC \to \nT$ mutation
    via the independent-sites model,
    or by a $\nC\nG \to \nT\nG$ mutation;
    in such cases
    the total instantaneous rate at which ACG changes to ATG is the sum of the rates
    (in this example, $\gamma + m_{\nC\nT}$).
    % Although this means there are many possible parameterizations,
    % note that the model is identifiable, not overparameterized.

    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{glemin2015quantification},
    which is effectively a selective pressure against G or C bases.
    We model the evolution of a \emph{single} sequence, not a population of sequences,
    imagining this sequence to be the consensus sequence of the population.
    A new mutation with an effect $s$ on fitness that occurs
    in a population of effective size $N_e$ becomes ubiquitous, rather than dying out,
    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
    If the mutation occurs at rate $\mu$ per individual, the total rate it occurs at in the population is $\mu N_e$.
    GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
    This is incorporated as follows: set the energy of pattern of pattern \nC{} or \nG{} to be $s$,
    so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
    Also, assume a $\phi$ function expressed in terms of an energy difference $e$:
    \begin{align*}
        \phi(e) = (1-\exp(-2e))/(1-\exp(-2eN)).
    \end{align*}
    Thus, if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
    then $\mu \, N_e \phi(E(y)-E(x))$ is the rate at which the mutation
    appears and successfully takes over in the population.

@@

\begin{example}[Gibbs sampling of the Ising model]
    In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
    imagined as a string of $n$ magnetic dipoles,
    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_{i=1}^{n-1} x_i x_{i+1} - \frac{1}{2} \gamma \sum_{i=1}^n x_i$.
    Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
    The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
    The following process, also known as ``Glauber dynamics'', preserves the stationary distribution:
    each site, independently at rate $\lambda$,
    forgets its spin,
    and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+\exp(-(E(x)-E(y)))$,
    this is:
    \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
          $+$  &   $-$   &  $\lambda$ \\
          $-$  &   $+$   &  $\lambda$
        \end{tabular}
        \qquad and \qquad
        \begin{tabular}{cc}
        pattern  &  energy \\
        $+-$ or $-+$  &   $\beta$ \\
        $+$ &   $\gamma$
        \end{tabular}
    \end{center}
    and
    \begin{align*}
        \phi(e) = \frac{1}{1+\exp(-e)} .
    \end{align*}

    This process has been studied by XXX.
\end{example}

\paragraph{The generator matrix}
We now describe concretely how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x \in \calS^n$ jumps to state $y \in \calS^n$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le n$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h,\ \text{and} \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e., if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
and for those positions $x$ matches with $u$ while $y$ matches with $v$.
Let $\calT$ be the set of these transition triples.

As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, i.e.,
\[
    J(i,x,y) = \{ (\mu,u,v) \in \calT \st x \xrightarrow{i,u,v} y \}.
\]
The rate ${G(n)}_{x,y}$ is then the sum of all matching transition rates,
multiplied by the fitness term,
namely
\begin{align} \label{eqn:G_defn}
    {G(n)}_{x,y} = \phi\left(E(y)-E(x)\right) \, \sum_{i=1}^n \sum_{(\mu, u, v) \in J(i,x,y)}  \mu ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
Indeed, we anticipate this matrix to be sparse because for most pairs $(x,y)$ there are no transition triples that can modify $x$ to make $y$.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
or ``uniformization'' \citep{hobolth2009simulation}.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu^*$ per site,
and then resolving each possible change in temporal order.
(Note that these ``possible changes'' are different than the ``proposed changes'' described above
in terms of the energy model.)
The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
    \mu^* = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
\]
Now sample possible changes:
in a sequence of length $L$, over a time period $t$,
first draw the number of possible changes, $N$,
from a Poisson distribution with mean $\mu^* t L$,
and then choose the times $s_k$ and locations $i_k$
at which possible changes occur uniformly
over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
Order the possible change times, so that
\[
    0 < s_1 < \cdots < s_N < t .
\]
(Note that this is equivalent to saying that possible
changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
Now, suppose we have determined the state just before time $s_k$ to be $X(s_k) = x$.
Then, a new state is then chosen at position $i_k$
with probability proportional to the corresponding rate, i.e.,
out of the possible transitions $\calT$, the $j^\text{th}$ one is chosen
with probability
\[
q_j = \begin{cases}
    % TODO peter, fix notation, please.
    \mu^{(j)}/\mu^* \qquad & \text{if } x_i^{(|u^{(j)}|)} = u^{(j)}  \\
    0 \qquad & \text{otherwise.}
\end{cases}
\]
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)
If triple $(\mu^{(j)}, u^{(j)}, v^{(j)})$ is chosen, then
the substring of $x$ matching $u^{(j)}$ that begins at position $i$
is replaced with $v^j$ to produce $X(s_k)$.

For instance, to simulate the TASEP, one has only to let $\mu^*=\lambda$, and at each $c_{ij}$
check if at site $i$ there is a 1 followed by a 0,
and if so, switch them.

@@

The problem at hand is to infer the parameters of the model based on the final state of the model after it has evolved from some known initial state.
% If the sequence is long enough and not too much time has gone by
% this should be feasible -- there should be enough information in our observations to do this reliably.
% If the time is too large, the process may ``saturate'' --
% if there are many changes at most sites, then in at least most models,
% we will lose most information about the dynamics,
% retaining information only about the parameters that affect the stationary distribution.
% In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
% while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
% In the CpG model it is not immediately clear what information is retained.
How can we extract the information?
These are Markov processes, on the state space $\calS^n$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible even for moderately sized $n$.
We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, this ignores dependencies between neighboring blocks,
so it is not clear how good this approximation might be.
% Intuitively, we want to increase the size of the context
% -- and the likelihood thus obtained is asymptotically correct for $m$ large,
% but computationally, we are stuck with small values of $m$.

@@

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\calP}{\mathcal{P}}  % set of selection pairs
\newcommand{\calM}{\mathcal{M}}  % set of potential mutations
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\nS}{\mbox{S}} % C or G
\newcommand{\nW}{\mbox{W}} % A or T
\newcommand{\nR}{\mbox{R}} % purine, A or G
\newcommand{\nY}{\mbox{Y}} % pyrimidine, C or T
\newcommand{\join}{\oplus}  % concatenation
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}
\newcommand{\like}{\mathcal L}
\newcommand{\loglike}{\ell}
\newcommand{\alike}{\widetilde{\like}}
\newcommand{\aloglike}{\widetilde{\loglike}}

@@

Early models of DNA sequence mutation
assumed that each nucleotide was equally likely to mutate
into any other nucleotide \citep{jukes1969evolution}.
These were followed by others that incorporated differences
in the mutation process, such as higher transition to transversion ratio
\citep{hasegawa1985dating,felsenstein1996hidden}.
Although the rates of mutation of a nucleotide
are known to depend on not only the nucleotide itself
but also nearby sequence context \citep{seplyarskiy2021population},
it is difficult to take this context into account when performing inference.

This situation from biological sequence analysis
is an example of a more commonly encountered class of models in statistics:
a lattice of sites,
with each site taking one of a finite collection of possible states,
and whose stochastic, temporal evolution is Markov and governed by a set of local rules.
For instance, at each site may sit
a nucelotide whose mutation rate depends on physical properties determined by the nearby DNA sequence;
a cell whose infection status depends on the state of its neighbors;
or a particle whose spin is perturbed by external noise to a state whose distribution depends on the local energy configuration.
Practical use of such models often requires
inferring transition rules
based on observations of the system at several time points,
or of several states evolved along a tree from a single starting point.
This paper concerns scalable methods for doing inference under such models using observations.

@@

In the context of genomics,
it is well-known that certain short nucleotide sequences
are in many organisms much more, or less, abundant than expected by chance \citep{burge1992underrepresentation},
due to the combined effects of context dependence of the nucleotide mutation process,
selective constraints on the function of the sequence,
and other processes such as biased gene conversion \citep{duret2009biased,arbeithuber2015crossovers}.
Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
is highly context dependent \citep{schaibley2013influence,gao2019overlooked,rodriguezgalindo2020germline}
and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution,herediagenestar2020extreme},
which provide quite different cellular contexts but can share commonalities due to shared molecular machinery.
In general, substitution rates depend on sequence context due to the interaction of a great many factors including:
(a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{goodman2002errorprone} or crossing over in meiosis \citep{arbeithuber2015crossovers});
(b) activity levels and structural properties of the enzymes responsible for repair
(e.g., Y-family polymerases \citep{goodman2013translesion,sale2012yfamily});
and (c) any functional effects of an error (e.g., lethal mutations will not be seen in a population census).
We are concerned with effects that are \emph{homogeneous} across the sequence,
so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.
Some of these effects -- particularly, activity levels of different enzymes in the germline --
will likely vary over evolutionary time,
and so comparisons between widely separated species will identify time-averaged mutation rates.
Studies have used a variety of exploratory techniques to identify sets of mutational patterns
co-occurring in comparisons (a) between different cancer types \citep{alexandrov2013signatures,shiraishi2015simple} and
(b) between different human populations \citep{harris2015evidence,harris2016rapid,mathieson2017differences},
to disentangle distinct signals presumably coming from the action of distinct sources of mutation.

@@

To formalize this notion of local dependence,
first define a \textit{pattern} to mean a contiguous sequence of states,
and let $|u|$ denote the length of a pattern $u$.
Given a state $x \in \calS$,
let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
% Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
% This is natural in many situations, but in others we might want to other boundary conditions,
% modeling extant, but unobserved, neighbors.
% We will mention this when necessary.
The dynamics of our stochastic process
are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$,
written as ``$u \to v$ at rate $\mu$''.
A model of context-dependent mutation
is defined by a set of all transition triples,
which we denote $\calT$.

Although our main focus is on nucleotides,
it is informative to consider other, simpler, models as well.
Here is one such example.

\begin{example}[TASEP]
  In the \emph{Totally Asymmetric Simple Exclusion Process}
  describes a set of particles moving among some sites,
  with at most one particle per site.
  Each site is either ``empty'' or ``occupied''
  (0 or 1, respectively),
  and each particle, independently at rate $\lambda$,
  checks whether the site to the right is empty,
  and if so, moves there.
  This process therefore has only one transition triple:

@@

  \noindent
  This only has one parameter.
  If we allow particles to exit from the right, and new particles to enter from the left,
  then from observing only starting and ending configurations
  we cannot tell with certainty which particles have moved where,
  so it is not obvious how to estimate the speed.
  However, summing over possible particle movements would allow us to
  compute the likelihood of a given configuration change,
  and so estimate the speed by maximum likelihood
  given an initial sequence, a final sequence and an elapsed time.

@@

To allow more compact specification of models,
we also introduce a ``potential'':
suppose that there is a collection of patterns $\{p_i\}$
such that each occurrence of $p_i$ adds a quantity $e_i$ to the total ``energy'' of a sequence,
and that the rate at which each possible change occurs is modulated by a function of the energy difference the change would produce.
This is natural if transition triples describe \emph{proposed} changes,
and that the probability a proposed change actually occurs depends on how much it affects the energy.

@@

\begin{example}[Genomic GC content]
    A genome sequence can be written using A, C, G, and T;
    in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
    Furthermore, it is a well-known observation that in many species,
    adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
    than either single nucleotide change under the independent model.
    Embellishing the single-nucleotide model with this additional rate results in the model defined by
    \begin{center}
      \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
        $x$  &  $y$  &  $m_{xy}$ & for $x, y \in \{\nA,\nC,\nG,\nT\}$ and $x \neq y$ \\
        \nC\nG   &  \nT\nG   &  $\gamma$ & \\
        \nC\nG   &  \nC\nA   &  $\gamma$ &
      \end{tabular},
    \end{center}
    where $\gamma$ is the additional CpG rate above the base mutation rate.
    Note that for a given pair of sequences,
    a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a $\nC \to \nT$ mutation
    via the independent-sites model,
    or by a $\nC\nG \to \nT\nG$ mutation;
    in such cases
    the total instantaneous rate at which ACG changes to ATG is the sum of the rates
    (in this example, $\gamma + m_{\nC\nT}$).
    % Although this means there are many possible parameterizations,
    % note that the model is identifiable, not overparameterized.

    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{glemin2015quantification},
    which is effectively a selective pressure against G or C bases.
    We model the evolution of a \emph{single} sequence, not a population of sequences,
    imagining this sequence to be the consensus sequence of the population.
    A new mutation with an effect $s$ on fitness that occurs
    in a population of effective size $N_e$ becomes ubiquitous, rather than dying out,
    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
    If the mutation occurs at rate $\mu$ per individual,
    the total rate it occurs at in the population is $\mu N_e$.
    GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
    This is incorporated as follows: set the energy of pattern of pattern \nC{} or \nG{} to be $s$,
    so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
    Also, define $\phi(e)$ to be the expected rate of fixations of ``selected'' mutations
    with advantage $e$:
    \begin{align*}
        \phi(e) = N_e (1-\exp(-2e))/(1-\exp(-2eN)).
    \end{align*}
    With these definitions,
    if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
    then $\mu \, \phi(E(y)-E(x))$ is the rate at which the mutation
    appears and successfully takes over in the population.

@@

\begin{example}[Gibbs sampling of the Ising model]
    In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
    imagined as a string of $n$ magnetic dipoles,
    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_{i=1}^{n-1} x_i x_{i+1} - \frac{1}{2} \gamma \sum_{i=1}^n x_i$.
    Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
    The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
    The following process, also known as ``Glauber dynamics'', preserves the stationary distribution:
    each site, independently at rate $\lambda$,
    forgets its spin,
    and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+\exp(-(E(x)-E(y)))$,
    this is:
    \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
          $+$  &   $-$   &  $\lambda$ \\
          $-$  &   $+$   &  $\lambda$
        \end{tabular}
        \qquad and \qquad
        \begin{tabular}{cc}
        pattern  &  energy \\
        $+-$ or $-+$  &   $\beta$ \\
        $+$ &   $\gamma$
        \end{tabular}
    \end{center}
    and
    \begin{align*}
        \phi(e) = \frac{1}{1+\exp(-e)} .
    \end{align*}
\end{example}

\paragraph{The generator matrix}
We now describe concretely how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x \in \calS^n$ jumps to state $y \in \calS^n$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le n$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h,\ \text{and} \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e., if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
and for those positions $x$ matches with $u$ while $y$ matches with $v$.

As before, let $\calT$ be the set of transition triples for the model.
As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, i.e.,
\[
    J(i,x,y) = \{ (\mu,u,v) \in \calT \st x \xrightarrow{i,u,v} y \}.
\]
The rate ${G(n)}_{x,y}$ is then the sum of all transition rates that can take $x$ to $y$,
multiplied by the fitness term,
namely
\begin{align} \label{eqn:G_defn}
    {G(n)}_{x,y} = \phi\left(E(y)-E(x)\right) \, \sum_{i=1}^n \sum_{(\mu, u, v) \in J(i,x,y)}  \mu ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
To make this the generator matrix for a Markov process,
we also set each $G(n)_{x,x}$ so that rows sum to zero.
Indeed, this matrix is very sparse because for most pairs $(x,y)$ there are no transition triples that can change $x$ into $y$.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
or ``uniformization'' \citep{hobolth2009simulation}.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu^*$ per site,
and then resolving each possible change in temporal order.
(Note that these ``possible changes'' are different than the ``proposed changes'' described above
in terms of the energy model.)
The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
    \mu^* 
    % = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
    = \max_x \sum_{y \neq x} G(L)_{x,y} .
\]
Note that we do not actually need to construct $G(L)$ to find the maximum rate.
Now sample possible changes:
in a sequence of length $L$, over a time period $t$,
first draw the number of possible changes, $N$,
from a Poisson distribution with mean $\mu^* t L$,
and then choose the times $s_k$ and locations $i_k$
at which possible changes occur uniformly
over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
Order the possible change times, so that
\[
    0 < s_1 < \cdots < s_N < t .
\]
(Note that this is equivalent to saying that possible
changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
Now, suppose we have determined the state just before time $s_k$ to be $X(s_k) = x$.
Then, a new state is then chosen at position $i_k$
with probability proportional to the corresponding rate, i.e.,
out of the possible transitions $\calT$, transition triple $(\mu(j), u(j), v(j))$ is chosen
with probability
\[
q_j = \begin{cases}
    \mu(j)/\mu^* \qquad & \text{if } x_i^{(|u(j)|)} = u(j)  \\
    0 \qquad & \text{otherwise.}
\end{cases}
\]
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0 \ge 0$.)
If triple $(\mu(j), u(j), v(j))$ is chosen, then
the substring of $x$ matching $u(j)$ that begins at position $i_k$
is replaced with $v(j)$ to produce $X(s_k)$.

For instance, to simulate the TASEP, one has only to let $\mu^*=\lambda$,
and at each possible change
check if at site $i_k$ there is a 1 followed by a 0,
and if so, switch them.

@@

The problem at hand is to infer the parameters of the model based on the final state of the model after it has evolved from some known initial state.
% If the sequence is long enough and not too much time has gone by
% this should be feasible -- there should be enough information in our observations to do this reliably.
% If the time is too large, the process may ``saturate'' --
% if there are many changes at most sites, then in at least most models,
% we will lose most information about the dynamics,
% retaining information only about the parameters that affect the stationary distribution.
% In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
% while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
% In the CpG model it is not immediately clear what information is retained.
How can we extract the information?
These are Markov processes, on the state space $\calS^L$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^L \times |\calS|^L$ matrix $G_L$ is clearly infeasible even for moderately sized $L$.
We can, however, compute \eqref{eqn:full_likelihood} with smaller $G_n$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, this ignores dependencies between neighboring blocks,
so it is not clear how good this approximation might be.
% Intuitively, we want to increase the size of the context
% -- and the likelihood thus obtained is asymptotically correct for $m$ large,
% but computationally, we are stuck with small values of $m$.

@@

%EM TODO grant acknowledgements
